Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-29 Manuscript/manuscript.Rmd
edits

We first sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically.

@@

We then investigated the proportion of class imbalance, defined as the ratio of TAD boundary regions versus the sum of both TAD and non-TAD boundary regions. We compared the estimated the number of TAD boundary regions ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) for each cell line, at each of the four resolutions of Hi-C data. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to modeling.

@@

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.

@@

In order to identify the optimal set of parameters necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of random forest classifiers built on different combinations of predictor types and re-sampling techniques (Figure 4). Each system in the framework was assumed to be composed of a random forest model constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

>>>>

We first sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were found to be specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically. This indicates a necessity for computationally specific models in predicting TAD boundary formations.

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.

@@

[include number of cell type specific features and number of features in common]

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->
